---
title: "Analysing steps per day"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

```


This is a document for the Peer Assessmentof the Reproducible Research course from coursera.

```{r, echo=TRUE}

data <-  read.csv(file = "activity.csv", stringsAsFactors = FALSE)
original_data <- data
```

We got only the complete cases

```{r, echo=TRUE}
data <- data[complete.cases(data),]
data$date <- as.Date(data$date)
```


Afeter processing data with dplyr we can show the total steps per day:

```{r, echo=TRUE}
data_total_steps <- select(data, steps, date) %>%
                  group_by(date) %>%
                  summarise(sum_steps_day = sum(steps) )

data_total_steps
```

We can show our findings with two similars plots. The frist one is the bar plot where we use a catergorical data to show our findings. The second one is the histogram, we´ll use a continous data to represent our findings with it:

```{r, echo=TRUE}
ggplot(data=data_total_steps, aes(x=date, y=sum_steps_day, fill=sum_steps_day)) + 
    geom_histogram(stat="identity") +
    xlab("Date") + ylab("Total Steps") +
    ggtitle("Total Steps by Day Histogram")
```

We calculate the mean and the median for the data.

```{r, echo=TRUE}


data_mean_median_steps <- select(data, steps, date) %>%
                            group_by(date) %>%
                            summarise(mean_steps_day = mean(steps), median_steps_day = median(steps) )

      
data_mean_median_steps
```

Now we´ll prepare the data to calculate the avarage number of steps grouped by 5-minute interval:

```{r, echo=TRUE}
data_5_minutes_interval <- group_by(data, interval) %>%
                            summarise(mean_steps = mean(steps))

data_5_minutes_interval
```                         

Let´s just put it into a time series plot.

```{r, echo=TRUE}

ggplot(data_5_minutes_interval, aes(x = mean_steps, y = interval)) +
    geom_line() +
    xlab("Mean Steps") + ylab("Interval") +
    ggtitle("Time Series: Mean steps in 5 min Interval")


```

Until now we used only the complete cases, but let´s do something different.

Let´s calculate the total missing values:

```{r, echo=TRUE}

missing_values <- sum(!complete.cases(original_data))
missing_values
```

We choosed to fill those lines with the mean of total steps

```{r, echo=TRUE}

mean_steps <- mean(data$steps)

original_data[!complete.cases(original_data), 1] <- mean_steps

```

Let´s so calculate the new value of total steps:

```{r}
original_data_total_steps <- select(original_data, steps, date) %>%
                  group_by(date) %>%
                  summarise(sum_steps_day = sum(steps) )

original_data_total_steps
```


Now let´s see it in a histogram:

```{r, echo=TRUE}
ggplot(data=original_data_total_steps, aes(x=date, y=sum_steps_day, fill=sum_steps_day)) + 
    geom_histogram(stat="identity") +
    xlab("Date") + ylab("Total Steps") +
    ggtitle("Total Steps by Day Histogram")
```

And calculate the mean and the median for the data.

```{r, echo=TRUE}


original_data_mean_median_steps <- select(original_data, steps, date) %>%
                            group_by(date) %>%
                            summarise(mean_steps_day = mean(steps), median_steps_day = median(steps) )

      
original_data_mean_median_steps
```

As we can see there were missing blocs in the first histogram that now were filled.

Now let´s see f thereare differences between weekends and weekdays.

We´ll now create a new table with one more variables, that will be a factor with two levels "weekday" and "weekend", lets name it type_day.

As I´m brazlian, the weekend days are named "sábado" and "domingo", equivalent to "saturday" and "sunday".

```{r, echo=TRUE}


original_data_weekdays <- mutate(original_data, 
                                          type_day = ifelse(weekdays(as.Date(date)) == "sábado" 
                                                            | weekdays(as.Date(date)) == "domingo", 
                                                            "weekend", "weekday")) %>%
                                  mutate(type_day = as.factor(type_day))
                                


      
```

Now let´s make a time series plot of the mean steps taken in determined interval on weekends and weekdays:

```{r, echo=TRUE}
original_data_5_min_weekdays <- group_by(original_data_weekdays, interval, type_day) %>%
                                summarise(mean_steps = mean(steps))


ggplot(original_data_5_min_weekdays, aes(x = mean_steps, y = interval, fill = type_day, color = type_day)) +
    geom_line() +
    xlab("Mean Steps") + ylab("Interval") +
    ggtitle("Time Series: Mean steps in 5 min Interval on Weekends and Weekdays")
```

